---
title: "Tree Maintenance Assignments"
format:
  html:
    embed-resources: true
    page-layout: full
execute:
  echo: false
  message: false
  warning: false
---

```{=html}
<style>
.quarto-layout-row { column-gap: 2em; }
</style>
```

```{r}
#| label: setup
#| include: false
source(here::here('Trees/Maintenance/AssignPeopleToTrees.R'))

trees_voronoi          = segment_voronoi(trees, people)
trees_nearest          = assign_capacitated_nearest(trees, people, tolerance = 0.3) |>
  swap_assignments(people)
trees_optimal          = assign_optimal_transport(trees, people)
trees_voronoi_balanced = assign_voronoi_rebalanced(trees, people)
trees_voronoi_weighted = assign_weighted_voronoi(trees, people)
trees_draft            = assign_draft(trees, people)
trees_clustered        = assign_geographic_clusters(trees, people)
```

::: {.panel-tabset}

## Comparison

:::{style="min-height: 5em;"}
All algorithms ranked by CV, total distance, and max single-tree distance.
Lower rank sum is better. Score combines the three metrics normalized by the
Voronoi (nearest-person) baseline.
:::

```{r}
#| label: comparison
all_assignments = list(
  'Nearest Person'   = trees_voronoi,
  'Balanced Nearest' = trees_nearest,
  'Min Distance'     = trees_optimal,
  'Rebalanced'       = trees_voronoi_balanced,
  'Weighted Zones'   = trees_voronoi_weighted,
  'Draft Pick'       = trees_draft,
  'Geographic'       = trees_clustered
)
compare_assignments(all_assignments, people)
```

## Nearest Person

:::{style="min-height: 5em;"}
Each tree is assigned to the closest volunteer. This is the simplest approach
and keeps travel distances short, but some volunteers may end up with many more
trees than others.
:::

```{r}
#| label: voronoi
#| layout: [[45, 55]]
summarize_assignments(trees_voronoi, people, 'Nearest Person')
show_assignments(trees_voronoi, people)
```

## Balanced Nearest

:::{style="min-height: 5em;"}
Trees are assigned to the closest volunteer who still has room. Volunteers near
more trees fill up first, then nearby trees go to the next-closest volunteer.
This spreads the workload more evenly while still keeping assignments fairly
close to each volunteer.
:::

```{r}
#| label: nearest
#| layout: [[45, 55]]
summarize_assignments(trees_nearest, people, 'Balanced Nearest')
show_assignments(trees_nearest, people)
```

## Minimum Distance

:::{style="min-height: 5em;"}
Assignments are chosen to minimize the total travel distance across all
volunteers while keeping everyone's workload roughly equal. This finds the
best overall balance between short trips and fair distribution.
:::

```{r}
#| label: optimal
#| layout: [[45, 55]]
summarize_assignments(trees_optimal, people, 'Minimum Distance')
show_assignments(trees_optimal, people)
```

## Rebalanced

:::{style="min-height: 5em;"}
Starts with the nearest-person assignment, then moves trees between neighboring
volunteers to even out the workload. Only trees near the boundary between two
volunteers are moved, so the assignments stay geographically sensible.
:::

```{r}
#| label: rebalanced
#| layout: [[45, 55]]
summarize_assignments(trees_voronoi_balanced, people, 'Rebalanced')
show_assignments(trees_voronoi_balanced, people)
```

## Weighted Zones

:::{style="min-height: 5em;"}
Adjusts the size of each volunteer's territory until everyone has roughly the
same number of trees. Volunteers with too few trees get a larger zone;
volunteers with too many get a smaller one. This keeps each person's trees in
a single contiguous area.
:::

```{r}
#| label: weighted
#| layout: [[45, 55]]
summarize_assignments(trees_voronoi_weighted, people, 'Weighted Zones')
show_assignments(trees_voronoi_weighted, people)
```

## Draft Pick

:::{style="min-height: 5em;"}
Volunteers take turns picking their nearest available tree. The volunteer
furthest below their fair share always picks next. This guarantees every
volunteer receives trees, even those who are geographically far from others.
:::

```{r}
#| label: draft
#| layout: [[45, 55]]
summarize_assignments(trees_draft, people, 'Draft Pick')
show_assignments(trees_draft, people)
```

## Geographic

:::{style="min-height: 5em;"}
Trees are grouped into geographically compact clusters using balanced k-means,
ignoring volunteer locations entirely. The number of clusters equals the total
number of volunteers. Each cluster is then assigned to the nearest volunteer,
with volunteers who represent multiple people receiving multiple clusters.
:::

```{r}
#| label: clustered
#| layout: [[45, 55]]
summarize_assignments(trees_clustered, people, 'Geographic')
show_assignments(trees_clustered, people)
```

:::
