---
title: "Tree Maintenance Assignments"
format:
  html:
    embed-resources: true
execute:
  echo: false
  message: false
  warning: false
---

```{r}
#| label: setup
#| include: false
source(here::here('Trees/AssignPeopleToTrees.R'))
```

::: {.panel-tabset}

## Nearest Person

Each tree is assigned to the closest volunteer. This is the simplest approach
and keeps travel distances short, but some volunteers may end up with many more
trees than others.

```{r}
#| label: voronoi
summarize_assignments(trees_voronoi, people, 'Nearest Person')
show_assignments(trees_voronoi, people)
```

## Balanced Nearest

Trees are assigned to the closest volunteer who still has room. Volunteers near
more trees fill up first, then nearby trees go to the next-closest volunteer.
This spreads the workload more evenly while still keeping assignments fairly
close to each volunteer.

```{r}
#| label: nearest
summarize_assignments(trees_nearest, people, 'Balanced Nearest')
show_assignments(trees_nearest, people)
```

## Minimum Distance

Assignments are chosen to minimize the total travel distance across all
volunteers while keeping everyone's workload roughly equal. This finds the
best overall balance between short trips and fair distribution.

```{r}
#| label: optimal
summarize_assignments(trees_optimal, people, 'Minimum Distance')
show_assignments(trees_optimal, people)
```

## Rebalanced

Starts with the nearest-person assignment, then moves trees between neighboring
volunteers to even out the workload. Only trees near the boundary between two
volunteers are moved, so the assignments stay geographically sensible.

```{r}
#| label: rebalanced
summarize_assignments(trees_voronoi_balanced, people, 'Rebalanced')
show_assignments(trees_voronoi_balanced, people)
```

## Weighted Zones

Adjusts the size of each volunteer's territory until everyone has roughly the
same number of trees. Volunteers with too few trees get a larger zone;
volunteers with too many get a smaller one. This keeps each person's trees in
a single contiguous area.

```{r}
#| label: weighted
summarize_assignments(trees_voronoi_weighted, people, 'Weighted Zones')
show_assignments(trees_voronoi_weighted, people)
```

:::
